%  acknow.tex

\section*{Acknowledgements}

I'd like to thank God for me giving the opportunity to get a PhD. His plans are always so much greater than even my wildest dreams. Randy Pausch, in his Last Lecture said, ``luck is where preparation meets opportunity''. This is a very insightful quote. I like to think and live my life by an  analogous definition, that of grace. Grace is where faith meets God's blessings. Grace is also defined as getting what you don't deserve. I feel that I am given tremendous blessings every day and thank God that He is so gracious to me.

I wouldn't have made it this far in my PhD endeavors if it wasn't for my wife, Patty. She is my everything and helps me so much each day. She has been by my side helping me with all of the little details of the whole process. It truly was a journey, one that I am so grateful that we have been able to take together. Not only is this the end of one chapter of our lives, but it is the beginning of the next.

I have a wonderful family and great friends. They have been so supportive of me in everything that I do. I know that they are so proud of me and my accomplishments, but I'm glad that I have their love and support. They make changing the world worth all of the work. 

My advisor, Jeanna Matthews, has been a true inspiration to me. Her passion for making a difference, helping people, and  her philosophy of ``leaving things better than you found them'' have made a significant impact on my life in many ways. I am thankful for her endless patience, guidance, and wisdom. I also thank her for helping me develop such a strong interest in networking, systems, and open source.

The Applied Computer Science Labs at Clarkson, specifically the Clarkson Open Source Institute (COSI) and the Internet Teaching Laboratory (ITL), have also been another wonderful inspiration and help to me. I have met so many people in the labs and developed so many great relationships over the years. I appreciate the many conversations and feedback that I have received from them over the years on my research, life, and the world in general.

I would like to thank my PhD committee for their thorough and insightful feedback on my dissertation. Also thanks to the many reviewers that read early drafts of chapters, paragraphs, and ideas. The conversations and feedback that I have received have turned this work into something that I could not have come up with all by myself.

I'd like to thank IBM for their generosity to higher education and specifically for funding me for two years with consecutive IBM PhD Fellowships. My mentors during the program, Sean Dague and Rick Harper, are still influential and impressive to me today. I'm so fortunate to have met both of them.

I know that by listing specific people I will inevitably leave someone out. However, I would like to acknowledge by name several close friends and colleagues who I have met during my PhD career. These guys have worked with me in one way or another over the years and they deserve recognition. A big thanks to Eli M. Dow, Wenjin Hu, Patrick F. Wilbur, Jim Owens, and Tao Yang. You guys are next for getting your PhDs. I pass that torch onto you. Best of luck. 

Last, but certainly not least, I would be remiss to forget to thank the open source community for producing great free software, helping me understand things, and having patience to help answer even the most trivial of questions. I hope that I can somehow give back to a free and open source software (FOSS) community that is very deserving of my efforts and thanks.



